---
title: 'Reproducible Reports - Project #2'
author: "Mike G"
date: "November 10, 2015"
output: html_document
---

# Title

### Questions:
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

columns: FATALIES, INJURIES


2. Across the United States, which types of events have the greatest economic consequences?

columns: PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, WFO

## Synopsis

## Data Processing  

### Load Data into R  

```{r, echo=TRUE}

# load storm data into a dataframe
storm <- read.table('repdata-data-StormData.csv.bz2', stringsAsFactors = FALSE,sep=",",header=TRUE)

# list data type for each column
str(storm)

```

### Data Processing - Population Health  

```{r}
# get events where injuries or fatalities occurred
require(dplyr)

people <- storm %>%
            select(EVTYPE, FATALITIES, INJURIES) %>%
            filter(FATALITIES > 0 | INJURIES > 0)

# look at type of events that impacted people's health
table(people$EVTYPE)

# We have over 200 events, so let's consolidate them for analysis.
mapEvent <- function(evt) {
    result <- "OTHER"
    evt <- toupper(evt)
    
    if (grepl('TORNADO',evt)) {
        result = 'TORNADO'
    }
    else if (grepl('FLOOD',evt) || grepl('RAIN',evt)) {
         result = 'FLOOD'
    }
    else if (grepl('SEA',evt) || grepl('COASTAL',evt) || grepl('CURRENT',evt)) {
         result = 'FLOOD'
    }
     else if (grepl('SURGE',evt) || grepl('SURF',evt) || grepl('RIVER',evt)) {
         result = 'FLOOD'
    }
    else if (grepl('THUNDERSTORM',evt) || grepl('HAIL',evt)) {
         result = 'THUNDERSTORM'
    }
    else if (grepl('SNOW',evt) || grepl('BLIZZARD',evt)) {
         result = 'SNOW'
    }
    else if (grepl('COLD',evt) || grepl('ICE',evt) || grepl('ICY',evt)) {
         result = 'LOW-TEMP'
    }
    else if (grepl('WINTER',evt)) {
         result = 'LOW-TEMP'
    }
    else if (grepl('HEAT',evt) || grepl('WARM',evt)) {
        result = 'HIGH-TEMP' 
    }
    else if (grepl('FIRE',evt)) {
        result = 'FIRE'
    }
    else if (grepl('HURRICANE',evt)) {
        result = 'HURRICANE'
    }
    else if (grepl('WIND',evt)) {
        result = 'WIND'
    }
    else if (grepl('LIGHTNING',evt)) {
        result = 'LIGHTNING'
    }
    else if (grepl('AVALANCHE',evt) || grepl('SLIDE',evt)) {
        result <- 'LANDSLIDE'
    }
    else if (grepl('DUST',evt)) {
        result <- 'DUST'
    }
    
    return(result)
}

# use consolidated event types for people dataframe
people$EVTYPE <- sapply(people$EVTYPE, mapEvent)

# convert event type to a factor
people$EVTYPE <- as.factor(people$EVTYPE)

```


### Data Processing - Economic Impact 

```{r}

# get events where crops or property damage occurred
property <- storm %>%
                select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
                filter(PROPDMG > 0 | CROPDMG > 0)

```

#### Common Unit for Economic Damage  
In the storm dataset, columns CROPDMGEXP and PROPDMGEXP contain the units for
economic damage listed in columns CROPDMG AND PROPDMG. The columns use the 
following symbols:  
* k or K - dollars in thousands
* m or M - dollars in millions
* b or B - dollars in billions  

For our analysis, we will convert all damage amounts to units of thousand of 
dollars.

```{r}

# convert units to upper case 
property$PROPDMGEXP <- toupper(property$PROPDMGEXP)
property$CROPDMGEXP <- toupper(property$CROPDMGEXP)

# remove rows with invalid values for units
values <- c('K','M','B')

property <- property %>%
            filter(PROPDMGEXP %in% values | CROPDMGEXP %in% values)

# use consolidated events for property dataframe
property$EVTYPE <- sapply(property$EVTYPE, mapEvent)

require(reshape2)

# collapse PROPDMGEXP and CROPDMGEXP into single column
property2 <- melt(property, id=c('EVTYPE','PROPDMG','CROPDMG'))
head(property2)

# collapse PROPDMG and CROPDMG into single column
property2 <- melt(property2, id=c('EVTYPE','variable','value'))
head(property2)

# remove column 2, since it is no longer useful for analysis
property2 <- property2[,-2]
head(property2)

# update names for column
names(property2) <- c('EVTYPE','UNIT','TYPE','AMOUNT')

str(property2)


convertUnits <- function(df) {
    # get data with units already in thousands 
    x1 <- df %>% filter(UNIT == 'K')
    x1$AMOUNT <- x1$AMOUNT * 10^-3
    
    # convert data with units of millions to thousands
    x2 <- df %>% filter(UNIT == 'M')
    #x2$AMOUNT <- x2$AMOUNT * 10^3
    
    # convert data with inits of billions to thousands
    x3 <- df %>% filter(UNIT == 'B')
    x3$AMOUNT <- x3$AMOUNT * 10^3
    
    # combine results
    result <- rbind(x1,x2,x3)
    result <- result[,c(1,3:4)]
    return(result)
}

# update dataset to use common unit of thousand of dollars
property3 <- convertUnits(property2)

# convert event type to factor
property3$EVTYPE <- as.factor(property3$EVTYPE)


```

## Results

### Events Impact on Population Health
```{r}
str(people)

# get number of injuries or fatalities per event
df <- people %>%
        group_by(EVTYPE) %>%
        summarise_each(funs(sum))

head(df)

# put data in long format
df <- melt(df, id=c('EVTYPE'))
names(df)

# update names for 2nd and 3rd columns
names(df)[2:3] <- c('TYPE','COUNT')

# draw plot to display results
require(ggplot2)

ggplot(df) + 
    geom_bar(aes(x=EVTYPE,y=COUNT,fill=TYPE), stat="identity") +
    ggtitle("Number of Fatalities & Injuries\n per Event (1950 - 2011)") +
    labs(x="",y="") +
    coord_flip()


```

### Economic Consequences of Events Impact
```{r}
str(property3)

df2 <- property3 %>%
        group_by(EVTYPE, TYPE) %>%
        summarise(AMOUNT=sum(AMOUNT))

ggplot(df2) +
    geom_bar(aes(x=EVTYPE,y=AMOUNT,fill=TYPE),stat="identity") +
    ggtitle("Value of Crop & Property\n Damage per Event (1950 - 2011") +
    labs(x="",y="Cost in millions") +
    coord_flip()

ggplot(property3) +
    geom_bar(aes(x=EVTYPE,y=AMOUNT,fill=TYPE),stat="identity") +
    ggtitle("Value of Crop & Property\n Damage per Event (1950 - 2011)") +
    labs(x="",y="Cost in millions") +
    coord_flip() +
    facet_wrap(~ TYPE)

```

