# Title: Storm Data Analysis


## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities across the country. To view the economic and health impact of adverse weather in the United States, I analyzed storm data for the years 1950 to 2011 from the US National Oceanic and Atmospheric Administration (NOA) storm database. 

Before analyzing the data, I first loaded it into a R dataframe. After loading the data, I created two dataframes, one that contains public health impact and another that contains economic impact. I then cleaned data by removing unnecessary rows, consolidating the number of events and converting economic cost to a common uint. 

Once the data was cleaned-up, I performed summary analysis on each event type to see which ones had the biggest impact on public health and economic impact. Once the summary data was generated, graphs were created to show the visual impact of each event type on public health and economic impact.  




## Data Processing  

### STEP 1: Load Data into R  

```{r, cache=TRUE}

# load storm data into a dataframe
storm <- read.table('repdata-data-StormData.csv.bz2', stringsAsFactors = FALSE,sep=",",header=TRUE)

# list data type for each column
str(storm)

```

### STEP 2: Data Processing - Population Health  
This section contains the steps executed to prepare storm data that had an impact on public health.  

```{r, cache=TRUE}
# get events where injuries or fatalities occurred
require(dplyr)

people <- storm %>%
            select(EVTYPE, FATALITIES, INJURIES) %>%
            filter(FATALITIES > 0 | INJURIES > 0)

# look at type of events that impacted people's health
table(people$EVTYPE)

# We have over 200 events, so let's create a function to consolidate them for analysis.
mapEvent <- function(evt) {
    result <- "OTHER"
    evt <- toupper(evt)
    
    if (grepl('TORNADO',evt)) {
        result = 'TORNADO'
    }
    else if (grepl('FLOOD',evt) || grepl('RAIN',evt)) {
         result = 'FLOOD'
    }
    else if (grepl('SEA',evt) || grepl('COASTAL',evt) || grepl('CURRENT',evt)) {
         result = 'FLOOD'
    }
     else if (grepl('SURGE',evt) || grepl('SURF',evt) || grepl('RIVER',evt)) {
         result = 'FLOOD'
    }
    else if (grepl('THUNDERSTORM',evt) || grepl('HAIL',evt)) {
         result = 'THUNDERSTORM'
    }
    else if (grepl('SNOW',evt) || grepl('BLIZZARD',evt)) {
         result = 'SNOW'
    }
    else if (grepl('COLD',evt) || grepl('ICE',evt) || grepl('ICY',evt)) {
         result = 'LOW-TEMP'
    }
    else if (grepl('WINTER',evt)) {
         result = 'LOW-TEMP'
    }
    else if (grepl('HEAT',evt) || grepl('WARM',evt)) {
        result = 'HIGH-TEMP' 
    }
    else if (grepl('FIRE',evt)) {
        result = 'FIRE'
    }
    else if (grepl('HURRICANE',evt)) {
        result = 'HURRICANE'
    }
    else if (grepl('WIND',evt)) {
        result = 'WIND'
    }
    else if (grepl('LIGHTNING',evt)) {
        result = 'LIGHTNING'
    }
    else if (grepl('AVALANCHE',evt) || grepl('SLIDE',evt)) {
        result <- 'LANDSLIDE'
    }
    else if (grepl('DUST',evt)) {
        result <- 'DUST'
    }
    
    return(result)
}

# use consolidated event types for people dataframe
people$EVTYPE <- sapply(people$EVTYPE, mapEvent)

# convert event type to a factor
people$EVTYPE <- as.factor(people$EVTYPE)

```


### STEP 3: Data Processing - Economic Impact 
This section contains the steps executed to prepare storm data that had an economic impact.  

```{r}

# get events where crops or property damage occurred
property <- storm %>%
                select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
                filter(PROPDMG > 0 | CROPDMG > 0)

```

#### Common Unit for Economic Damage  
In the storm dataset, columns CROPDMGEXP and PROPDMGEXP contain the units for economic damage listed in columns CROPDMG AND PROPDMG. The columns use the following symbols:  
* k or K - dollars in thousands
* m or M - dollars in millions
* b or B - dollars in billions  

For our analysis, we will convert all units to millions of dollars.

```{r}

# convert units to upper case 
property$PROPDMGEXP <- toupper(property$PROPDMGEXP)
property$CROPDMGEXP <- toupper(property$CROPDMGEXP)

# remove rows with invalid values for units
values <- c('K','M','B')

property <- property %>%
            filter(PROPDMGEXP %in% values | CROPDMGEXP %in% values)


require(reshape2)

# collapse PROPDMGEXP and CROPDMGEXP into single column
property2 <- melt(property, id=c('EVTYPE','PROPDMG','CROPDMG'))
head(property2)

# collapse PROPDMG and CROPDMG into single column
property2 <- melt(property2, id=c('EVTYPE','variable','value'))
head(property2)

# remove column 2, since it is no longer useful for analysis
property2 <- property2[,-2]
head(property2)

# update names for column
names(property2)[2:4] <- c('UNIT','TYPE','AMOUNT')

str(property2)

# function used to convert damage cost to millions of dollars
convertUnits <- function(df) {
    # convert data with units of thousands to millions 
    x1 <- df %>% filter(UNIT == 'K')
    x1$AMOUNT <- x1$AMOUNT * 10^-3
    
    # no conversion for data with units already in millions
    x2 <- df %>% filter(UNIT == 'M')
    
    # convert data with inits of billions to millions
    x3 <- df %>% filter(UNIT == 'B')
    x3$AMOUNT <- x3$AMOUNT * 10^3
    
    # combine results
    result <- rbind(x1,x2,x3)
    result <- result[,c(1,3:4)]
    return(result)
}

# update dataset to use units of millions for damage amount
property3 <- convertUnits(property2)

# use consolidated events for property dataframe
property3$EVTYPE <- sapply(property3$EVTYPE, mapEvent)

# convert event type to factor
property3$EVTYPE <- as.factor(property3$EVTYPE)

```

## Results  
Now that the data is the desired state, we can perform the analysis required to answer the following questions:  

1. Across the United State, which types of events are most harmful with respect to population health?  
2. Across the United States, which types of events have the greatest economic consequences?  

### Part A: Events Impact on Population Health
```{r}
str(people)

# get number of injuries or fatalities per event
df <- people %>%
        group_by(EVTYPE) %>%
        summarise_each(funs(sum))

head(df)

# put data in long format
df <- melt(df, id=c('EVTYPE'))
names(df)

# update names for 2nd and 3rd columns
names(df)[2:3] <- c('TYPE','COUNT')

# draw plot to display results
require(ggplot2)
require(gridExtra)

# draw graph showing absolute values
g1 <- ggplot(df) + 
    geom_bar(aes(x=EVTYPE,y=COUNT,fill=TYPE), 
                    stat="identity") +
    ggtitle("Number of Fatalities & Injuries\n per Event (1950 - 2011)") +
    labs(x="",y="") +
    coord_flip()

# convert count to log10 to show more details
g2 <- ggplot(df) + 
    geom_bar(aes(x=EVTYPE,y=log10(COUNT),fill=TYPE), 
                    stat="identity") +
    ggtitle("Number of Fatalities & Injuries\n per Event (1950 - 2011)") +
    labs(x="",y="Units in log10") +
    coord_flip()

# plot graphs
grid.arrange(g1, g2, ncol=1)

```

### Part B: Economic Consequences of Events Impact
```{r}
str(property3)

df2 <- property3 %>%
        group_by(EVTYPE, TYPE) %>%
        summarise(AMOUNT=sum(AMOUNT))

g3 <- ggplot(df2) +
    geom_bar(aes(x=EVTYPE,y=AMOUNT,fill=TYPE),stat="identity") +
    ggtitle("Value of Crop & Property\n Damage per Event (1950 - 2011") +
    labs(x="",y="Cost in millions") +
    coord_flip()

g4 <- ggplot(df2) +
    geom_bar(aes(x=EVTYPE,y=log10(AMOUNT),fill=TYPE),stat="identity") +
    ggtitle("Value of Crop & Property\n Damage per Event (1950 - 2011)") +
    labs(x="",y="log10 Cost in millions") +
    coord_flip() 
    
    grid.arrange(g3, g4, ncol=1)

```

